A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay

### 2021-05-21 Paper/NatureAstronomy/main_resubmission2.tex

% Required packages
\usepackage{caption}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{xspace}
\usepackage{graphicx}

@@

%%%  SELECT DRAFT VERSION (CHANGETEXT) OR FINAL 
%\usepackage[draft,xcolor=dvipsnames]{changes}
%\linenumbers
\usepackage[final,xcolor=dvipsnames]{changes}

@@

\replaced[id=ed]{
When the light from a distant stellar explosion passes very near to a foreground galaxy or cluster, gravitational lensing can cause it to appear as multiple images on the sky.  Such strongly-lensed supernovae can be used to constrain the cosmic expansion rate \cite{refsdal_possibility_1964} and dark energy models \cite{treu_time_2016}.  Achieving these cosmological goals requires many more lensed supernova discoveries and more efficient observational follow-up.   Here we report the discovery of a multiply-imaged supernova that will enable a time delay measurement with an uncertainty of $<1\%$.  It appeared in an evolved galaxy at $z=1.95$, gravitationally lensed by a massive foreground galaxy cluster. It is likely a Type Ia supernova---the explosion of a low-mass stellar remnant, whose light curve can be used to measure cosmic distances.  In archival Hubble Space Telescope imaging, three lensed images of the supernova are detected with relative time delays of $<$200 days.  We predict a fourth image will appear close to the cluster core in the year 2037$\pm$2.   Observation and modeling of that future image could provide a time delay precision of $\approx 7$ days over an extraordinary 20 year baseline.  This object demonstrates that cosmologically useful time delays can be measured with minimal observational cost using cluster-lensed supernovae.}{
  We report the discovery of a strongly-lensed supernova (SN) explosion that will enable a time delay measurement with an uncertainty of $<1\%$. This unprecedented scenario will enable a rare and highly precise measurement of the Hubble\added[id=r1]{-LeMa\^{i}tre} constant from a single lensed SN.  The SN is in an evolved galaxy at $z=1.95$, gravitationally lensed by the massive foreground galaxy cluster MACS J0138.0-2155.  This is the most distant multiply-imaged supernova yet discovered.  We classify the supernova as \added[id=r2]{most likely} a Type Ia \deleted[id=r2]{with {$>90\%$} probability} based on archival imaging observations from the \textit{Hubble Space Telescope}.  Three lensed images of the SN are detected with relative time delays of $<$200 days, and we predict that a fourth image in the component closest to the cluster core will appear in the year 2037$\pm$2.  Observing the light curve of that future image could provide a time delay precision of $\approx 7$ days over an extraordinary 20 year baseline. 
}
\end{sciabstract}

@@

Observations of Type Ia supernova (SN) explosions have played a key role in mapping the cosmic expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate 
\cite{riess_observational_1998,perlmutter_measurements_1999} %Riess_large_2019}.  
Determining the nature of dark energy and how it may evolve over time is a primary goal for the large-scale cosmology experiments of the 2020s
\cite{amendola_cosmology_2013,spergel_wide_2015,Ivezic_lsst_2019}.
\deleted[id=ed]{A series of} Recent investigations into the expansion rate of the universe (the Hubble-LeMa\^itre constant; $H_0$) have found that measurements from the local universe are significantly different from the value inferred from measurement of the cosmic microwave background \deleted[id=ed]{(CMB)} radiation \cite{Riess_large_2019,aghanim_planck_2018}.  Resolving this tension in $H_0$ measurements may reveal new physics from the early universe, and to do so requires multiple 
independent cosmological probes \cite{verde_tensions_2019}.

One promising cosmological tool uses %time delays of 
gravitational lens systems in which a background
source appears as multiple images that arrive to the observer with relative delays \cite{einstein_lens_1936}.
%, which %for a given lens potential 
When such a strongly-lensed source is variable, one can measure the time delay between any pair of lensed images and derive a ratio of angular diameter distances to the foreground lens and background source. 
This distance ratio is sensitive to cosmological parameters, such as 
%the expansion rate of the universe, 
$H_0$ \cite{refsdal_possibility_1964} and the dark energy equation-of-state, $w$ \cite{coe_cosmological_2009}.%,linder_lensing_2011}.
In recent years this method has been successfully applied to lensed
quasars %cite{suyu_cosmology_2014,bonvin_cosmograil_2019}, 
with six high-precision measurements to date \cite{wong_h0licow_2019}. As this sample of time delay lenses grows to several dozen, it is expected to deliver a measurement of $H_0$ with 1\% precision \cite{suyu_cosmological_2018}. \added[id=r1]{A future sample with $\sim100$ well-measured lensing time delays will be a competitive tool for dark energy studies \cite{treu_time_2016}.}
%\cite{linder_lensing_2011,jee_time-delay_2016,suyu_HOLISMOKES_2020}.}

Gravitationally lensed SN with multiple images present an attractive new addition to this field. Most notably, SN exhibit relatively simple photometric behavior, with well-understood light curve shapes and colors---in contrast to the stochastic variation of quasars.  However, \deleted[id=ed]{despite the promise of lensed SN time delay cosmography as a cosmological distance probe,} to date there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_sn_2016}. The second, {\it SN 2016geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2019}. \added[id=r1]{An earlier discovery of an unusually luminous SN \cite{chornock_ps1-10afx_2013} was also shown to be a strongly-lensed SNIa,  \cite{quimby_detection_2014} %quimby_extraordinary_2013,
though the multiple images were not resolved.}  Here we present the discovery 
of a third lensed SN resolved into multiple images: \SNABC.

We discovered \SNABC ({\it AstroNotes citation})
using data from the Hubble Space Telescope (HST) program {\it REsolved QUIEscent Magnified Galaxies} (REQUIEM, HST-GO-15663, PI:Akhshik). This project targets rare examples of massive galaxies with low specific star formation rates that have been magnified by strong gravitational lensing. %These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular galaxy targeted by REQUIEM
is MRG-M0138, first discovered as a massive red galaxy (MRG) at $z=1.95$ \cite{newman_resolving_2018} behind the massive galaxy cluster MACS J0138.0-2155 \cite{ebeling_macs_2001}.
MRG-M0138 is quadruply lensed by a foreground galaxy cluster at $z=0.338$.  
During analysis of REQUIEM observations obtained 13--14 July 2019
%\cite{materials_methods_2020} 
we discovered three point 
sources that were present in archival HST images from 18--19 July 2016, part of the program (HST-GO-14496; PI:Newman)
that first confirmed the MRG-M0138 galaxy as a strongly-lensed object 
\added[id=ed]{(Methods: Observations)}. 
Each point source is within 5 arcseconds of one of the four MRG-M0138 images.  None of the
three point sources are present in the REQUIEM HST data in 2019 (Fig. \ref{fig:layout}). We infer that 
the point sources are multiple images of a single astrophysical 
transient in MRG-M0138, most likely a SN. %\cite{materials_methods_2020}.  
%The official designation for this multiply-imaged astrophysical transient is \TNSname. 
%Formal designation as a SN requires spectroscopic confirmation, but with a high-confidence photometric classification we coin the name \SNABC for this unusual source.

To construct a lens model for the MACS J0138.0-2155 cluster we use the \lenstool software \cite{jullo_bayesian_2007,kneib_lenstool_2011} to model the mass distribution in the cluster core as the combination of a cluster-scale and multiple galaxy-scale potentials 
\added[id=ed]{(Methods: Lens Modeling)}.
%\cite{materials_methods_2020}.  
To avoid unintended bias, we kept the lens model development completely separate from the analysis of the SN.  Only upon \deleted[id=ed]{independent} completion of both were the results combined for the analysis described here.  The input model constraints are the positions and redshifts of the MRG-M0138 galaxy at $z=1.95$ (both the \replaced[id=r3]{galaxy's centroid position }{central light peak} and the SN location in each image) as well as a multiply-imaged background 
galaxy at $z=0.766$, both having secure spectroscopic redshifts
\added[id=ed]{(Methods: Observations)}.
%\cite{materials_methods_2020}.  
From this model we derive estimates for the lensing magnification and time delay of each of the SN images, including the predicted fourth image (Table \ref{tab:time_delays}).
The lensing model predicts that the SN should appear in the fourth MRG-M0138 image in the year 2037$\pm$\replaced[id=r1]{2}{1.5}, demagnified with $\mu=0.4\pm0.2$. The predicted location for this future image is $\alpha=$01:38:04.15, $\delta=-$21:55:24.659.
A fifth image will also appear at a still later date, located near the center of the cluster and much more significantly demagnified, so it will not be easily observable.  \added[id=r1]{We anticipate that future lens modeling of the cluster may improve on these predictions} \added[id=ed]{(Supplementary Note: Future Work}).

To realize the cosmological promise of \SNABC 
we need to estimate the age of each SN image, which in turn constrains the lensing time delays. For this
goal, it is helpful to have a firm determination of the transient’s class. Expected time delays and magnifications from the lens model exclude any of the various rapidly evolving and low-luminosity stellar transient classes, strongly suggesting that it is a SN. The first-order SN distinction remaining is between a Type Ia SN---the explosion of a white dwarf star in a binary system---and a core collapse SN (CCSN)---the end-point of a star with mass $>10 M_{\odot}$. 
The properties of the host galaxy can inform this classification because CCSN are limited to galaxies with young stellar populations. Limits on the specific star formation rate and age for this host, MRG-M0138, show it to be a massive but very quiescent and evolved galaxy, %($\log(M_*/M_\odot)=11.7\pm0.2; )$ , 
unlikely to retain any significant population of high-mass stars \cite{newman_resolving_2018}. Based on observed properties of the host galaxy alone, we find a 62-75\% probability that \SNABC is a Type Ia SN. Adopting the lens model magnifications for the three observed SN images (Table \ref{tab:time_delays}) we can locate each SN image in color-magnitude space (Fig. \ref{fig:class}). \replaced[id=r1]{After magnification correction, all three images are still brighter than expected for a Type Ia SN, which may indicate that a lens-modeling degeneracy is at play \cite{falco_model-dependent_1985, schneider_source-position_2014}. Nevertheless, the magnification-corrected \SNABC data are more consistent with the Type Ia population than any CC SN sub-class (Fig.~\ref{fig:class}a,b), and demonstrate the expected evolution of a Type Ia SN color and brightness over $\sim 100$ days (Fig.~\ref{fig:class}c).}{This provides more classification leverage, yielding p(Ia)=95\%.}
By also including the model-predicted time delays, we can treat the three SN images effectively as three points on a common SN light curve, and we find p(Ia)=94\% 
\added[id=ed]{(Methods: Classification)}.
%\cite{materials_methods_2020}. 
\replaced[id=r2]{An improved classification can be achieved }{These final classifications are strongly conditional on the priors, and we look forward to an improved classification} with spectroscopy and multi-band photometry upon arrival of the fourth image.  \added[id=r2]{For the remainder of this analysis, we proceed under the assumption that \SNABC is indeed a Type Ia SN---though the subsequent analysis could be revised to achieve similar results with a different underlying SN model.}

The color of a SNIa evolves substantially over its lifetime as the photosphere expands and cools, revealing different layers of the expanding shell and driving episodes of recombination \cite{kasen_type_2009}. %woosley_type_2007
Since the phenomenon of gravitational lensing in general is achromatic, this color evolution makes it possible to derive an age constraint that is largely independent of the lens model. %\cite{materials_methods_2020}.  
Combining this information with magnification constraints from the lens modeling helps break parameter degeneracies, yielding the measured delays in Table \ref{tab:time_delays}
\added[id=ed]{(Methods: Time Delay)}.
%\cite{materials_methods_2020}. 
Remarkably, the ages of image 1 and 3 are constrained to better than $\pm$20 days, despite having only a single epoch of photometric data.    These uncertainties may be 
further reduced when the future fourth image is observed with high-precision, multi-epoch photometry.  Such a light curve will pin down the intrinsic SN light curve parameters that are shared by all images, and break remaining parameter degeneracies.
%\added[id=ed]{(Supplementary Note: Future Observations)}.

\added[id=r1]{As with any single time-delay system, \SNABC will not be particularly powerful as a solitary cosmological probe.  However,} future large-scale surveys such as the Vera C. Rubin Observatory 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their mission lifetimes, \added[id=r1]{though the vast majority of these will be lensed by galaxy-scale deflectors \cite{oguri_gravitationally_2010,  goldstein_rates_2019, pierel_projected_2020}.} %wojtak_magnified_2019, 
\deleted[id=r1]{However, many of these lensed SN will require substantial follow-up resources to be made cosmologically useful.} Thus, almost all will have significantly shorter 
delays (of order 10--100 days) and therefore higher relative uncertainties than \SNABC.  Since it is the {\it fractional} time delay uncertainty that 
propagates through to any time delay distance measurement, the extraordinarily long time delays of cluster-lensed SNe like \SNABC can deliver significantly better time delay precision, with comparable observational cost.  In fact, the long time baseline of lensed SN like \SNABC effectively insures that their cosmological precision is not limited by time delay measurement uncertainty. 
\added[id=r1]{Cluster-scale lenses are more complex than galaxy-scale lenses, but they generally have ``independent'' measurements of the magnification from several multiply-imaged systems in the same field. As such, they can provide a valuable check on systematics related to specific lens modeling software, microlensing variation, and more.}

In Figure \ref{fig:cosmo} we illustrate the cosmological value of time delay cosmography with objects like \SNABC  by considering a small sample of 5 ``\reqlike'' SN\added[id=r1]{---with $<1\%$ time delay precision and the same redshift ratio.} 
For a comparison on equal footing, we evaluate equal-sized lensed SN samples 
that mimic those expected from the Rubin Observatory and  the Roman Space Telescope.  Differences in the 
ratio of source to lens redshift account for the different cosmological parameter covariances, reflected in the orientations of the contours in these plots. Because of the unusually large redshift ratio of \SNABC, the constraints 
from the \reqlike sample are nicely complementary to the low-$z$ Rubin Observatory and high-$z$ Roman Space Telescope samples. 
Furthermore, when we assume that the lens modeling uncertainty is equal for each sample (amounting to a 5\% uncertainty in each time delay distance), the \reqlike sample has leverage that is comparable to the same number of lensed SN from the wide-field surveys. However, if we improve  the lensing constraints to 2\% precision, the \reqlike contours tighten dramatically, reflecting the fact that these long-time-baseline events are limited only by the lens modeling uncertainty.

To achieve these cosmological constraints in practice, such a sample could be developed with regular monitoring of cluster-scale lenses, partnered with modest follow-up to characterize any lensed SN discovered
\added[id=ed]{(Supplementary Note: Future Work}).
%\cite{materials_methods_2020}.  
As \SNABC shows, even a single epoch of photometric data at the time of detection may be sufficient to provide the necessary anchor point for this promising cosmological tool.  
%After that, all we need is patience.

@@

\clearpage
\begin{figure*}
    \centering
    \includegraphics[draft=False,width=0.9\textwidth]{Paper/Figures/fig1_layout.pdf}
    \caption{Overview of the MACSJ0138 cluster field and the locations of the \SNABC images  (SN1--3) and its multiply-imaged host galaxy (H1--4). The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  
    \added[id=r1]{Labels marked 3.1--3.4 indicate the locations of a separate  multiply-imaged [O\textsc{ii}]-emitter at $z=0.77$ used to help constrain the lensing potential.} 
    Panels \emph{b}--\emph{i} show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels \emph{bcde} show the imaging from 2016 July where the SN was visible and \emph{fghi} show the later imaging from 2019 July where the the SN has faded away.  Panels \emph{aei} indicate the location of the fourth image predicted to appear in $\approx$2037 \added[id=r3]{and panel \emph{a} shows the final and highly demagnified fifth image (``SN5'')}. All panels use the late-epoch F125W imaging for the green channel; nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, which helps to constrain the relative age of each SN image and the transient classification as a likely Type Ia supernova explosion. }
    \label{fig:layout}
\end{figure*}

>>>>

% Required packages
\usepackage{caption}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{threeparttable}

@@

%%%  SELECT DRAFT VERSION (CHANGETEXT) OR FINAL 
\usepackage[draft,xcolor=dvipsnames]{changes}
\linenumbers
%\usepackage[final,xcolor=dvipsnames]{changes}

@@

  When the light from a distant stellar explosion passes very near to a foreground galaxy or cluster, gravitational lensing can cause it to appear as multiple images on the sky.  Such strongly-lensed supernovae can be used to constrain the cosmic expansion rate \cite{refsdal_possibility_1964} and dark energy models \cite{treu_time_2016}.  Achieving these cosmological goals \replaced[id=r2]{will require many lensed supernovae with precise time delay measurements.}{requires many more lensed supernova discoveries suitable for precise time delay measurement.}   Here we report the discovery of a multiply-imaged supernova that will enable a time delay measurement with an uncertainty of $<1\%$.  It appeared in an evolved galaxy at $z=1.95$, gravitationally lensed by a massive foreground galaxy cluster. It is likely a Type Ia supernova---the explosion of a low-mass stellar remnant, whose light curve can be used to measure cosmic distances.  In archival Hubble Space Telescope imaging, three lensed images of the supernova are detected with relative time delays of $<$200 days.  We predict a fourth image will appear close to the cluster core in the year 2037$\pm$2. 
\added[id=r1]{The SN classification and the predicted reappearance time could be improved with further lens modeling and a comprehensive analysis of systematic uncertainties.}  
  Observation \replaced[id=r1]{of the fourth }{and modeling of that future} image could provide a time delay precision of $\approx 7$ days over an extraordinary 20 year baseline. 
  \deleted[id=r2]{ This object demonstrates that cosmologically useful time delays can be measured with minimal observational cost using cluster-lensed supernovae.}

\end{sciabstract}

@@

Observations of Type Ia supernova (SN) explosions have played a key role in mapping the cosmic expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate 
\cite{riess_observational_1998,perlmutter_measurements_1999} %Riess_large_2019}.  
Determining the nature of dark energy and how it may evolve over time is a primary goal for the large-scale cosmology experiments of the 2020s
\cite{amendola_cosmology_2013,spergel_wide_2015,Ivezic_lsst_2019}.
 Recent investigations into the expansion rate of the universe (the Hubble-LeMa\^itre constant; $H_0$) have found that measurements from the local universe are significantly different from the value inferred from measurement of the cosmic microwave background radiation \cite{Riess_large_2019,aghanim_planck_2018}.  \replaced[id=r2]{
 The community is actively attempting to resolve this tension in $H_0$ measurements by mitigating systematic uncertainties  or revealing new physics from the early universe \cite{verde_tensions_2019}. Either resolution will require multiple independent cosmological probes.}{
Resolving this tension in $H_0$ measurements may reveal new physics from the early universe, and to do so requires multiple 
independent cosmological probes.}

One promising cosmological tool uses %time delays of 
gravitational lens systems in which a background
source appears as multiple images that arrive to the observer with relative delays \cite{einstein_lens_1936}.
%, which %for a given lens potential 
When such a strongly-lensed source is variable, one can measure the time delay between any pair of lensed images and derive a ratio of angular diameter distances to the foreground lens and background source. 
This distance ratio is sensitive to cosmological parameters, such as 
%the expansion rate of the universe, 
$H_0$ \cite{refsdal_possibility_1964} and the dark energy equation-of-state, $w$ \cite{holz_seeing_2001}. 
In recent years this method has been successfully applied to lensed
quasars %cite{suyu_cosmology_2014,bonvin_cosmograil_2019}, 
with \replaced[id=r3]{seven}{six} high-precision measurements to date \replaced[id=r3]{\cite{wong_h0licow_2019}}{\cite{birrer_tdcosmo_2020}}. As this sample of time delay lenses grows to several dozen, it is expected to deliver a measurement of $H_0$ with 1\% precision \cite{suyu_cosmological_2018}. A future sample with $\sim100$ well-measured lensing time delays will be a competitive tool for dark energy studies \cite{treu_time_2016}.
%\cite{linder_lensing_2011,jee_time-delay_2016,suyu_HOLISMOKES_2020}.}

Gravitationally lensed SN with multiple images 
\replaced[id=r2]{could become an important part of this time-delay cosmology sample that will develop over the next 10-20 years.}{
present an attractive new addition to this field. Most notably, SN exhibit}
\added[id=r2]{Lensed SN are attractive for this purpose because they have} relatively simple photometric behavior, with well-understood light curve shapes and colors---in contrast to the stochastic variation of quasars.  However,  to date there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_sn_2016}. The second, {\it SN 2016geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2019}. An earlier discovery of an unusually luminous SN \cite{chornock_ps1-10afx_2013} was also shown to be a strongly-lensed SNIa,  \cite{quimby_detection_2014} %quimby_extraordinary_2013,
though the multiple images were not resolved.  
Here we present the discovery 
of a third lensed SN resolved into multiple images: \SNABC.

We discovered \SNABC ({\it AstroNotes citation})
using data from the Hubble Space Telescope (HST) program {\it REsolved QUIEscent Magnified Galaxies} (REQUIEM, HST-GO-15663, PI:Akhshik). This project targets rare examples of massive galaxies with low specific star formation rates that have been magnified by strong gravitational lensing. %These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular galaxy targeted by REQUIEM
is MRG-M0138, first discovered as a massive red galaxy (MRG) at $z=1.95$ \cite{newman_resolving_2018} behind the massive galaxy cluster MACS J0138.0-2155 \cite{ebeling_macs_2001}.
MRG-M0138 is quadruply lensed by a foreground galaxy cluster at $z=0.338$.  
During analysis of REQUIEM observations obtained 13--14 July 2019
%\cite{materials_methods_2020} 
we discovered three point 
sources that were present in archival HST images from 18--19 July 2016, part of the program (HST-GO-14496; PI:Newman)
that first confirmed the MRG-M0138 galaxy as a strongly-lensed object 
(Methods: Observations). 
Each point source is within 5 arcseconds of one of the four MRG-M0138 images.  None of the
three point sources are present in the REQUIEM HST data in 2019 (Fig. \ref{fig:layout}). We infer that 
the point sources are multiple images of a single astrophysical 
transient in MRG-M0138, most likely a SN. %\cite{materials_methods_2020}.  
%The official designation for this multiply-imaged astrophysical transient is \TNSname. 
%Formal designation as a SN requires spectroscopic confirmation, but with a high-confidence photometric classification we coin the name \SNABC for this unusual source.

To construct a lens model for the MACS J0138.0-2155 cluster we use the \lenstool software \cite{jullo_bayesian_2007,kneib_lenstool_2011} to model the mass distribution in the cluster core as the combination of a cluster-scale and multiple galaxy-scale potentials 
(Methods: Lens Modeling).
%\cite{materials_methods_2020}.  
To avoid unintended bias, we kept the lens model development completely separate from the analysis of the SN.  Only upon  completion of both were the results combined for the analysis described here.  The input model constraints are the positions and redshifts of the MRG-M0138 galaxy at $z=1.95$ (both the galaxy's centroid position and the SN location in each image) as well as a multiply-imaged background 
galaxy at $z=0.766$, both having secure spectroscopic redshifts
(Methods: Observations).
%\cite{materials_methods_2020}.  
From this model we derive estimates for the lensing magnification and time delay of each of the SN images, including the predicted fourth image (Table \ref{tab:time_delays}).
The lensing model predicts that the SN should appear in the fourth MRG-M0138 image in the year 2037$\pm$2, demagnified with $\mu=0.4\pm0.2$. The predicted location for this future image is $\alpha=$01:38:04.15, $\delta=-$21:55:24.659.
A fifth image will also appear at a still later date, located near the center of the cluster and much more significantly demagnified, so it will not be easily observable.  We anticipate that future lens modeling of the cluster may improve on these predictions (Supplementary Note: Future Work). \added[id=r1]{For example, 
our analysis adopted only a single form for the density profiles, and did not incorporate constraints from stellar kinematics or pixel-level surface brightness data from the multiply-imaged systems.}

To realize the cosmological promise of \SNABC 
we need to estimate the age of each SN image, which in turn constrains the lensing time delays. For this
goal, it is helpful to have a firm determination of the transient’s class. Expected time delays and magnifications from the lens model exclude any of the various rapidly evolving and low-luminosity stellar transient classes, strongly suggesting that it is a SN. The first-order SN distinction remaining is between a Type Ia SN---the explosion of a white dwarf star in a binary system---and a core collapse SN (CCSN)---the end-point of a star with mass $>10 M_{\odot}$. 
The properties of the host galaxy can inform this classification because CCSN are limited to galaxies with young stellar populations. Limits on the specific star formation rate and age for this host, MRG-M0138, show it to be a massive but very quiescent and evolved galaxy, %($\log(M_*/M_\odot)=11.7\pm0.2; )$ , 
unlikely to retain any significant population of high-mass stars \cite{newman_resolving_2018}. Based on observed properties of the host galaxy alone, we find a 62-75\% probability that \SNABC is a Type Ia SN. Adopting the lens model magnifications for the three observed SN images (Table \ref{tab:time_delays}) we can locate each SN image in color-magnitude space (Fig. \ref{fig:class}). After magnification correction, all three images are still brighter than expected for a Type Ia SN, which may indicate that a lens-modeling degeneracy is at play \cite{falco_model-dependent_1985, schneider_source-position_2014}. Nevertheless, the magnification-corrected \SNABC data are more consistent with the Type Ia population than any CC SN sub-class (Fig.~\ref{fig:class}a,b), and demonstrate the expected evolution of a Type Ia SN color and brightness over $\sim 100$ days (Fig.~\ref{fig:class}c).
By also including the model-predicted time delays, we can treat the three SN images effectively as three points on a common SN light curve, and we find p(Ia)=94\% 
(Methods: Classification).
%\cite{materials_methods_2020}. 
An improved classification can be achieved  with spectroscopy and multi-band photometry upon arrival of the fourth image.  For the remainder of this analysis, we proceed under the assumption that \SNABC is indeed a Type Ia SN---though the subsequent analysis could be revised to achieve similar results with a different underlying SN model.

The color of a SNIa evolves substantially over its lifetime as the photosphere expands and cools, revealing different layers of the expanding shell and driving episodes of recombination \cite{kasen_type_2009}. %woosley_type_2007
Since the phenomenon of gravitational lensing in general is achromatic, this color evolution makes it possible to derive an age constraint that is largely independent of the lens model. %\cite{materials_methods_2020}.  
Combining this information with magnification constraints from the lens modeling helps break parameter degeneracies, yielding the measured delays in Table \ref{tab:time_delays}
(Methods: Time Delay).
%\cite{materials_methods_2020}. 
Remarkably, the ages of image 1 and 3 are constrained to better than $\pm$20 days, despite having only a single epoch of photometric data.    These uncertainties may be 
further reduced when the future fourth image is observed with high-precision, multi-epoch photometry.  Such a light curve will pin down the intrinsic SN light curve parameters that are shared by all images, and break remaining parameter degeneracies.
%(Supplementary Note: Future Observations).

As with any single time-delay system, \SNABC will not be particularly powerful as a solitary cosmological probe.  However, future large-scale surveys such as the Vera C. Rubin Observatory 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their mission lifetimes, though the vast majority of these will be lensed by galaxy-scale deflectors \cite{oguri_gravitationally_2010,  goldstein_rates_2019, pierel_projected_2020}. %wojtak_magnified_2019, 
 Thus, almost all will have significantly shorter 
delays (of order 10--100 days) and therefore higher relative uncertainties than \SNABC.  Since it is the {\it fractional} time delay uncertainty that 
propagates through to any time delay distance measurement, the extraordinarily long time delays of cluster-lensed SNe like \SNABC can deliver significantly better time delay precision, with comparable observational cost.  In fact, the long time baseline of lensed SN like \SNABC effectively insures that their cosmological precision is not limited by time delay measurement uncertainty. 
Cluster-scale lenses are more complex than galaxy-scale lenses, but they generally have ``independent'' measurements of the magnification from several multiply-imaged systems in the same field. As such, they can provide a valuable check on systematics related to specific lens modeling software, microlensing variation, and more.

\deleted{In Figure \ref{fig:cosmo} we illustrate the cosmological value of time delay cosmography with objects like \SNABC  by considering a small sample of 5 ``\reqlike'' SN---with $<1\%$ time delay precision and the same redshift ratio. 
For a comparison on equal footing, we evaluate equal-sized lensed SN samples 
that mimic those expected from the Rubin Observatory and  the Roman Space Telescope.  Differences in the 
ratio of source to lens redshift account for the different cosmological parameter covariances, reflected in the orientations of the contours in these plots. Because of the unusually large redshift ratio of \SNABC, the constraints 
from the \SNABC-like sample are nicely complementary to the low-$z$ Rubin Observatory and high-$z$ Roman Space Telescope samples. 
Furthermore, when we assume that the lens modeling uncertainty is equal for each sample (amounting to a 5\% uncertainty in each time delay distance), the \SNABC-like sample has leverage that is comparable to the same number of lensed SN from the wide-field surveys. However, if we improve  the lensing constraints to 2\% precision, the \SNABC-like contours tighten dramatically, reflecting the fact that these long-time-baseline events are limited only by the lens modeling uncertainty.}

\added[id=r2]{
The first multiply-imaged SN discovery, SN Refsdal, has shown that time delay cosmography with a cluster-lensed SN is viable \cite{rodney_sn_2016,kelly_deja_2016}.  However, the observing campaign for SN Refsdal was extraordinary, deploying more than 75 HST orbits over three years. Significant observational investment has been required for high-precision time delay measurement of lensed quasars, such as decade-long programs \cite{millon_cosmograil_2020} or daily monitoring for high-cadence light curves \cite{millon_tdcosmo_2020}.  In the case of \SNABC it will be possible to achieve similar time delay precision over the 20-year baseline with just a single imaging epoch as the anchor point. }

\replaced[id=r2]{A sample of \SNABC-like events}{To achieve these cosmological constraints in practice, such a sample} could be developed with regular monitoring of cluster-scale lenses, partnered with modest follow-up to characterize any lensed SN discovered
(Supplementary Note: Future Work). \deleted[id=r2]{As \SNABC shows, even a single epoch of photometric data at the time of detection may be sufficient to provide the necessary anchor point for this promising cosmological tool. }  \added[id=r3]{ Even with a concerted effort to discover  cluster-lensed SN, the vast majority of time delay cosmography constraints will always come from  quasars and SN lensed by {\em galaxy}-scale lenses. 
The systematic biases of cluster lens models are very different from those of galaxy scale lens modeling, so systems like \SNABC can be an asset for validating time delay cosmology systematics.
%We leave a detailed exploration of the cosmological impact of this and other cluster-lensed SN to future work. 
}

@@

\clearpage
\begin{figure*}
    \centering
    \includegraphics[draft=False,width=0.9\textwidth]{Paper/Figures/fig1_layout.pdf}
    \caption{Overview of the MACSJ0138 cluster field and the locations of the \SNABC images  (SN1--3) and its multiply-imaged host galaxy (H1--4). The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  
    Labels marked 3.1--3.4 indicate the locations of a separate  multiply-imaged [O\textsc{ii}]-emitter at $z=0.77$ used to help constrain the lensing potential. 
    Panels \emph{b}--\emph{i} show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels \emph{bcde} show the imaging from 2016 July where the SN was visible and \emph{fghi} show the later imaging from 2019 July where the the SN has faded away.  Panels \emph{aei} indicate the location of the fourth image predicted to appear in $\approx$2037 and panel \emph{a} shows the final and highly demagnified fifth image (``SN5''). All panels use the late-epoch F125W imaging for the green channel; nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, which helps to constrain the relative age of each SN image and the transient classification as a likely Type Ia supernova explosion. }
    \label{fig:layout}
\end{figure*}

\clearpage
\begin{figure}
  \centering
    \includegraphics[draft=False, width=\textwidth]{Paper/Figures/snRequiem_classcontours_timeline_hist.png}
    \caption{Classification information for \SNABC based on its position in color-magnitude space. (a) 
    The observed photometry for the three SN images is shown as open markers. Vertical dotted lines show the lens magnification corrections using our preferred lens model, with a filled grey marker indicating the corrected (demagnified) magnitude. Error bars indicate the observational and systematic uncertainty, including the range of alternative magnification corrections encompassed by lens model variations.
    Contours show the population distributions for normal SNe of Type Ia (red), Type Ib/Ic (gold), and Type II (green), drawn with two contour levels enclosing 68\%  and 95\%  of each SN population.  Each SN sub-class was simulated at $z=1.95$, and samples from their expected light curves were drawn uniformly in time.  
    (b) Marginalized distributions along the color dimension for the three SN sub-classes (using the same color scheme).  The simulated populations have been scaled according to the expected explosion rates in the SN host galaxy, based on its stellar population properties. Open markers show the observed colors again.
    (c) Zoomed-in view of the color-mag space marked by the grey box in panel a.
    The evolution of a typical Type Ia SN at $z=1.95$  is shown by a colored line, with the line color indicating SN age in observer-frame days relative to peak brightness.  White diamonds correspond to the times labeled on the colorbar below. Grey shading shows the typical range of luminosities and colors observed for the Type Ia SN population in the nearby universe. %\cite{wang_determination_2006}.
    Although the magnification-corrected data are brighter than expected for most SNIa, they are consistent both with the overall SNIa population, and with the SNIa color-mag vs time curve.  
    %An interactive 3-D version of this figure is available online at \href{https://plot.ly/~jpierel/6}{https://plot.ly/~jpierel/6}
    \label{fig:class}
    }
\end{figure}

%%% FIGURE 3 REMOVED 
%\clearpage
%\begin{figure}
%    \centering
%    \includegraphics[draft=False,trim={2cm, 3cm, 3cm, 4cm},clip,width=\textwidth]{Paper/Figures/snrequiem_hw_w0wa_apples_to_lsst_ngrst_4panel.pdf}
%    \caption{  
%    The sensitivity of \SNABC to the expansion rate of the universe ($H_0$) and dark energy equation-of-state parameters ($w_0, w_a$). Dashed lines and the black marker show the input values (upper right of the top panels) that were used for this simulation. Contours show the 1$\sigma$ confidence region for a projected sample of 5 lensed SNIa time delays from the Rubin Observatory (blue) and the Roman Space Telescope (red) compared to 5 events like \SNABC (black).  The left and right columns assume a 5\% and 2\% contribution to the distance uncertainty from modeling the lens potential, respectively. 
%    In order to isolate the plotted parameters, all contours are constructed assuming perfect knowledge of the cosmological parameters that are not shown. 
%    }
%    \label{fig:cosmo}
%\end{figure}
